---
title: "Mad4Hatter QC Report"
format:
  html:
    embed-resources: true
    code-fold: true
    toc: true
params:
  results_dir: "results"
  manifest_path: "manifest.tsv"
  output_dir: "QC_report"
  standardise_sample_name: true
  read_threshold: 100
  read_filter: 0
  af_filter: 0.01
  negative_control_read_threshold: 50
  reprep_threshold: 0.5
  repool_threshold: 0.75
  allele_col: "pseudocigar_masked"
  long_target_threshold: 275
  filter_wsaf_final_allele_table: true
  reaction_as_pool: false
execute:
  echo: false
  warning: false
  message: false
---

This report summarizes QC statistics assessing the success of a **Mad4hatter** run. See the
[workspace README](https://github.com/EPPIcenter/mad4hatter-wdl) for details on the required
inputs and QC summaries produced below.

Note: A target is considered successfully amplified if it has more than `read_threshold` reads,
where `read_threshold` is a threshold set as a workflow input.

```{r setup}
library(ggplot2)
library(plotly, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(tidyr)
library(ggbeeswarm)
library(stringr)
library(tibble)
library(purrr, quietly = TRUE, warn.conflicts = FALSE)
library(kableExtra, quietly = TRUE, warn.conflicts = FALSE)

results_dir <- params$results_dir
manifest_path <- params$manifest_path
output_dir <- params$output_dir
standardise_sample_name <- params$standardise_sample_name
read_threshold <- params$read_threshold
read_filter <- params$read_filter
af_filter <- params$af_filter
negative_control_read_threshold <- params$negative_control_read_threshold
reprep_threshold <- params$reprep_threshold
repool_threshold <- params$repool_threshold
allele_col <- params$allele_col
long_target_threshold <- params$long_target_threshold
filter_wsaf_final_allele_table <- params$filter_wsaf_final_allele_table
reaction_as_pool <- params$reaction_as_pool

# Path to outputs, all relative to output_dir once written
reprep_repool_summary_path <- "reprep_repool_summary.csv"
polyclonal_information_path <- "positive_control_polyclonal_info.csv"
neg_control_information_path <- "negative_control_amplified_targets.csv"
filtered_allele_data_path <- "allele_data_filtered.txt"
filtered_collapsed_allele_data_path <- "allele_data_collapsed_filtered.txt"
missing_targets_report_path <- "missing_targets_report.csv"
samples_to_repool_path <- "samples_to_repool.csv"
samples_to_reprep_path <- "samples_to_reprep.csv"
missing_samples_report_path <- "missing_samples_report.csv"
filtered_resmarker_table_path <- "resmarker_table_filtered.txt"
filtered_resmarker_microhaplotype_table_path <- "resmarker_microhaplotype_table_filtered.txt"
```

```{r load-manifest}
manifest <- read.csv(manifest_path, sep = "\t")
# Ensure manifest column is sample_name
colnames(manifest)[1] <- "sample_name"
```

```{r input-summary}
cat("============================================\n")
cat("USER CONFIGURATION SUMMARY\n")
cat("============================================\n")
cat("Results Directory:", results_dir, "\n")
cat("Manifest File:", manifest_path, "\n")
cat("Read Threshold for Successful Amplification:", read_threshold, "\n")
cat("Minimum Reads per ASV (Positive Control Filter):", read_filter, "\n")
cat("Minimum Allele Frequency per ASV (Positive Control Filter):", af_filter, "\n")
cat("Allele Column:", allele_col, "\n")
cat("============================================\n")
```

```{r load-data}
sample_coverage_path <- file.path(results_dir, "sample_coverage_postprocessed.txt")
amplicon_coverage_path <- file.path(results_dir, "amplicon_coverage_postprocessed.txt")
allele_data_path <- file.path(results_dir, "allele_data.txt")
collapsed_allele_data_path <- file.path(results_dir, "allele_data_collapsed.txt")
panel_information_path <- file.path(results_dir, "panel_information/amplicon_info.tsv")
resmarker_table_path <- file.path(results_dir, "resistance_marker_module/resmarker_table.txt")
resmarker_microhaplotype_table_path <- file.path(results_dir, "resistance_marker_module/resmarker_microhaplotype_table.txt")

sample_coverage <- read.csv(sample_coverage_path, sep = "\t")
amplicon_coverage <- read.delim(amplicon_coverage_path)
allele_data <- read.csv(allele_data_path, sep = "\t")
panel_information <- read.csv(panel_information_path, sep = "\t")

# Standardize sample names if requested
standardise_name <- function(sample_name) {
  if (standardise_sample_name) {
    sample_name %>%
      sub("_L001$", "", .) %>%
      word(., start = 1, end = -2, sep = "_")
  } else {
    sample_name
  }
}

amplicon_coverage <- amplicon_coverage %>% mutate(sample_name = standardise_name(sample_name))
allele_data <- allele_data %>% mutate(sample_name = standardise_name(sample_name))
sample_coverage <- sample_coverage %>% mutate(sample_name = standardise_name(sample_name))
```

## Section A: Locus Filtering

Here we filter out two types of targets for QC calculations only:

* **Targets amplifying non-Plasmodium falciparum species**, which are expected to amplify only if
  other species are present.
* **Long amplicons (>long_target_threshold bp, including primers)**, which tend to underperform
  due to their length.

These targets should not be considered when assessing the success of a sequencing run, but will
still be included in the final `allele_data_filtered.txt` output.

```{r locus-filtering}
# Find long targets to remove
panel_information <- panel_information %>%
  mutate(
    target_length = (insert_end - insert_start) +
      nchar(fwd_primer) +
      nchar(rev_primer)
  )

long_loci <- panel_information %>%
  filter(target_length > long_target_threshold) %>%
  pull(target_name)

cat("Long targets being removed:\n")
cat(unlist(long_loci), sep = "\n")

# Species loci to remove
old_species_loci_names <- c(
  "Pf3D7_13_v3-1041593-1041860-1AB",
  "PmUG01_12_v1-1397996-1398245-1AB",
  "PocGH01_12_v1-1106456-1106697-1AB",
  "PvP01_12_v1-1184983-1185208-1AB",
  "PKNH_12_v2-198869-199113-1AB"
)
new_species_loci_names <- c(
  "Pf3D7_13_v3-1041624-1041829",
  "PmUG01_12_v1-1398020-1398213",
  "PocGH01_12_v1-1106482-1106671",
  "PvP01_12_v1-1185007-1185184",
  "PKNH_12_v2-0198893-0199083"
)
M1addon_loci <- c(
  "PKNH_MIT_v2-0002298-0002488",
  "PmUG01_MIT_v1-0000606-0000784",
  "PocGH01_MIT_v2-0000358-0000550",
  "PvP01_MIT_v2-0004470-0004642"
)

excluded_loci <- unique(c(old_species_loci_names, new_species_loci_names, M1addon_loci, long_loci))

amplicon_coverage_qc <- amplicon_coverage[!amplicon_coverage$target_name %in% excluded_loci, ]
allele_data_qc <- allele_data[!allele_data$target_name %in% excluded_loci, ]
panel_information_qc <- panel_information[!panel_information$target_name %in% excluded_loci, ]

cat("\nTarget count summary:\n")
cat("  Total targets in amplicon_coverage:", n_distinct(amplicon_coverage$target_name), "\n")
cat("  Targets used for QC (after filtering):", n_distinct(amplicon_coverage_qc$target_name), "\n")
cat("  Excluded targets will still appear in final allele_data_filtered.txt output\n")
```

## Section B: Pool File Loading & Reaction Assignment

```{r reaction-assignment}
# reaction here refers to 1 of the 2 mPCR reactions.
if (reaction_as_pool) {
  panel_information_qc <- panel_information_qc %>% mutate(reaction = pool)
} else {
  panel_information_qc <- panel_information_qc %>% mutate(reaction = (str_extract(pool, "\\d+")))
}

# Count loci per reaction (only including loci that were not filtered above)
nloci_table <- panel_information_qc %>%
  count(reaction) %>%
  rename(nreactionloci = n)
print(nloci_table)

merge_data <- function(manifest, sample_coverage, standardise_sample_name) {
  sample_coverage %>%
    mutate(reads = suppressWarnings(as.numeric(reads))) %>%
    replace_na(list(reads = 0)) %>%
    pivot_wider(names_from = stage, values_from = reads) %>%
    inner_join(manifest, by = "sample_name")
}

# Reformat sample_coverage and add sample information from manifest
sample_coverage_with_manifest <- merge_data(manifest, sample_coverage, standardise_sample_name)

# Add manifest to amplicon coverage
amplicon_coverage_with_manifest <- amplicon_coverage_qc %>%
  inner_join(manifest, by = c("sample_name")) %>%
  left_join(panel_information_qc %>% select(target_name, reaction), by = "target_name")

# Create sample summary stats
summary_samples <- amplicon_coverage_with_manifest %>%
  left_join(nloci_table, by = "reaction") %>%
  group_by(sample_name, Batch, SampleType, reaction, nreactionloci) %>%
  summarize(
    reads_per_reaction = sum(OutputPostprocessing),
    n_good_loci = sum(OutputPostprocessing > read_threshold),
    .groups = "drop"
  ) %>%
  group_by(sample_name, Batch, SampleType) %>%
  mutate(reads_per_sample = sum(reads_per_reaction)) %>%
  ungroup() %>%
  mutate(prop_good_loci = n_good_loci / nreactionloci) %>%
  inner_join(manifest, by = join_by(sample_name, Batch, SampleType))

# Plotting setup
sample_colours <- c(
  "negative" = "red3",
  "positive" = "blue3",
  "sample" = "darkgrey",
  "empty" = "black"
)
reaction_colours <- c("orangered2", "skyblue", "black", "red", "green3", "blue", "cyan", "magenta", "yellow", "gray")

cat("Sample Summary\n")
cat(length(unique(manifest$sample_name)), "samples in manifest", "\n")
cat(length(unique(allele_data$sample_name)), "samples in allele data file", "\n")
cat(length(unique(allele_data_qc$sample_name)), "samples in filtered allele data file", "\n")
cat(length(unique(sample_coverage$sample_name)), "samples in manifest and sample coverage file", "\n")
cat(length(unique(amplicon_coverage_qc$sample_name)), "samples in manifest and amplicon coverage file", "\n")
```

## Plate Layout

The plate layout provides a visual representation of sample organization, including the placement
of positive and negative controls.

```{r plate-template}
create_tabsetOfGgplotObjects <- function(ggplotObjectsList) {
  cat("\n\n::: {.panel-tabset}\n\n")
  iwalk(ggplotObjectsList, ~ {
    cat("## ", .y, "\n\n")
    print(.x)
    cat("\n\n")
  })
  cat(":::\n")
}

create_plate_template <- function(summary_samples, batches, nrows = 8, ncols = 12) {
  expand.grid(
    Batch = batches,
    y = 1:nrows,
    Column = 1:ncols
  ) %>%
    mutate(
      ymin = y - 0.45,
      ymax = y + 0.45,
      xmin = Column - 0.45,
      xmax = Column + 0.45
    ) %>%
    mutate(Row = toupper(rev(letters[1:nrows])[y])) %>%
    left_join(
      summary_samples %>% select(Batch, Column, Row, SampleType) %>% distinct(),
      by = c("Batch", "Column", "Row" = "Row")
    )
}

quadrants <- create_plate_template(summary_samples, unique(summary_samples$Batch))

plot_plate <- function(quadrants_batch, sample_colours, batch_name) {
  ggplot(quadrants_batch) +
    geom_tile(aes(x = Column, y = y, fill = SampleType), color = "white") +
    scale_y_continuous(breaks = quadrants_batch$y, labels = quadrants_batch$Row) +
    xlab("Column") +
    ylab("Row") +
    scale_x_continuous(breaks = 1:12, labels = as.character(1:12)) +
    geom_rect(
      data = quadrants_batch,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      color = "black", fill = NA, linewidth = 1
    ) +
    scale_fill_manual(values = sample_colours) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 12),
      strip.text = element_text(size = 14, face = "bold"),
      panel.grid = element_blank(),
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    ) +
    guides(fill = guide_legend(override.aes = list(color = "black", size = 1))) +
    ggtitle(paste0("Plate Map: ", batch_name)) +
    coord_fixed(ratio = 0.75)
}

plot_plate_layouts <- function(plate_data, sample_colours) {
  plate_maps <- list()
  plate_data$SampleType[is.na(plate_data$SampleType)] <- "empty"
  for (b in unique(plate_data$Batch)) {
    quadrants_batch <- plate_data %>% filter(Batch == b)
    plate_maps[[b]] <- plot_plate(quadrants_batch, sample_colours, b)
  }
  plate_maps
}
```

```{r plate-layout}
#| results: asis
plate_maps <- plot_plate_layouts(quadrants, sample_colours)
cat(create_tabsetOfGgplotObjects(plate_maps))
```

## Primer Dimer Content

Here we visualise the proportion of sequencing reads classified as primer dimers.

```{r dimer-plot}
generate_dimer_plot <- function(sample_coverage_with_manifest, sample_colours) {
  ggplot(data = sample_coverage_with_manifest %>% arrange(desc(SampleType))) +
    geom_point(
      aes(x = Input + 0.9, y = (1 - OutputPostprocessing / Input) * 100, color = SampleType),
      shape = 1, alpha = 0.8, stroke = 1
    ) +
    scale_x_log10() +
    facet_wrap(~Batch) +
    ylab("% Dimers") +
    xlab("Input Reads") +
    ggtitle("Dimer Content") +
    scale_color_manual(values = sample_colours)
}

generate_dimer_plot(sample_coverage_with_manifest, sample_colours)
```

## Balancing Across Batches

Distribution of total reads per sample across batches.

```{r balancing-plot}
generate_balancing_plot <- function(summary_samples, sample_colours) {
  ggplot() +
    ggbeeswarm::geom_quasirandom(
      data = summary_samples %>%
        select(sample_name, Batch, reads_per_sample, SampleType) %>%
        distinct(),
      aes(x = Batch, y = reads_per_sample + 0.9, color = SampleType)
    ) +
    scale_y_log10() +
    scale_color_manual(values = sample_colours) +
    ylab("Total Reads for Sample") +
    ggtitle("Balancing Across Batches") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
}

generate_balancing_plot(summary_samples, sample_colours)
```

## Control Summary

### Multiallelic Targets in Positive Controls

```{r positive-control-polyclonality}
allele_data_qc <- allele_data_qc %>%
  group_by(sample_name, target_name) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()

pos_control_allele_data <- allele_data_qc %>%
  inner_join(manifest %>% filter(SampleType == "positive"), by = "sample_name")

if (allele_col == "pseudocigar_masked") {
  pos_control_data <- pos_control_allele_data %>%
    group_by(sample_name, target_name, pseudocigar_masked) %>%
    summarise(reads = sum(reads), AlleleFreq = sum(AlleleFreq), .groups = "drop")
} else {
  pos_control_data <- pos_control_allele_data
}

locus_summary <- pos_control_data %>%
  group_by(sample_name, target_name) %>%
  summarise(
    NumASVs_Meeting_Threshold = sum(reads > read_filter & AlleleFreq > af_filter),
    .groups = "drop"
  ) %>%
  mutate(
    Category = case_when(
      NumASVs_Meeting_Threshold == 1 ~ "Monoallelic",
      NumASVs_Meeting_Threshold == 2 ~ "2 Alleles",
      NumASVs_Meeting_Threshold > 2 ~ ">2 Alleles"
    )
  )
fill_colors <- c("Monoallelic" = "Gray", "2 Alleles" = "orangered2", ">2 Alleles" = "cyan")

result <- locus_summary %>%
  group_by(sample_name, Category) %>%
  summarise(Count = n(), .groups = "drop")

ggplot(result, aes(x = sample_name, y = Count, fill = Category, text = paste("Count:", Count))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Alleles per Target for Positive Controls", x = "Sample Name", y = "Number of Targets") +
  theme_minimal() +
  scale_fill_manual(values = fill_colors) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

polyclonal_information <- pos_control_data %>%
  inner_join(locus_summary, by = c("sample_name", "target_name")) %>%
  filter(Category %in% c("2 Alleles", ">2 Alleles"))

if (nrow(polyclonal_information) == 0) {
  cat("No multiallelic targets found for positive controls.\n")
} else {
  head(polyclonal_information)
}
```

### Negative Control Contamination

#### Read Summary per Negative Control

```{r negative-control-setup}
amplicons_negative <- amplicon_coverage_with_manifest %>%
  filter(SampleType == "negative") %>%
  mutate(negative = paste0(Batch, " Well: ", toupper(Row), Column))

batches_with_negatives <- amplicons_negative %>%
  group_by(Batch) %>%
  summarize(n_controls = n_distinct(negative), .groups = "drop") %>%
  arrange(Batch)

n_batches <- nrow(batches_with_negatives)

ncol_layout <- 3
base_width <- 12
base_height_per_row <- 2.5
min_height <- 6
max_height <- 20

cat("Negative controls by batch:\n")
print(batches_with_negatives)
cat("Total batches with negative controls:", n_batches, "\n")
```

```{r negative-control-histogram-fn}
plot_negative_control_histogram <- function(data, reaction_colours, batch_name = NULL, remove_zeros = TRUE) {
  data <- data %>% mutate(reaction = factor(reaction))

  if (remove_zeros) {
    data <- data %>% filter(!is.na(OutputPostprocessing) & OutputPostprocessing > 0)
  }
  if (nrow(data) == 0) {
    return(NULL)
  }

  max_read <- suppressWarnings(max(data$OutputPostprocessing, na.rm = TRUE))
  if (!is.finite(max_read)) max_read <- 0
  x_max <- if (max_read < 30) 40 else (max_read + 10)
  x_limits <- c(0, x_max)

  total_targets <- nrow(data)
  y_max <- if (total_targets < 90) 10 else (total_targets + 10)
  y_limits <- c(0.01, y_max)

  title_suffix <- if (!is.null(batch_name)) paste0(" - ", batch_name) else ""

  ggplot(data) +
    geom_histogram(aes(x = OutputPostprocessing, fill = reaction), bins = 12, boundary = 0) +
    facet_wrap(~negative, ncol = ncol_layout, scales = "fixed") +
    xlab("Reads") +
    ylab("Number of targets") +
    ggtitle(paste0("Read Distribution per Negative Control", title_suffix)) +
    scale_fill_manual(values = reaction_colours) +
    coord_cartesian(xlim = x_limits, ylim = y_limits, expand = FALSE) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text = element_text(size = 12),
      strip.text = element_text(size = 11, face = "bold"),
      panel.spacing = unit(1, "lines")
    )
}
```

```{r negative-control-histograms}
#| results: asis
if (n_batches == 1) {
  batch_data <- amplicons_negative
  total_negatives <- n_distinct(batch_data$negative)
  batch_data_nonzero <- batch_data %>% filter(!is.na(OutputPostprocessing) & OutputPostprocessing > 0)
  plotted_negatives <- n_distinct(batch_data_nonzero$negative)
  zero_read_negatives <- total_negatives - plotted_negatives

  if (nrow(batch_data_nonzero) == 0) {
    cat("No negative controls with reads in this batch — nothing to plot.\n")
  } else {
    cat(sprintf("Plotting %d/%d negatives (%d zero-read)\n\n", plotted_negatives, total_negatives, zero_read_negatives))
    n_controls <- plotted_negatives
    batch_height <- min(max_height, max(min_height, ceiling(n_controls / ncol_layout) * base_height_per_row))
    knitr::opts_chunk$set(fig.width = base_width, fig.height = batch_height)
    p <- plot_negative_control_histogram(batch_data_nonzero, reaction_colours)
    if (!is.null(p)) print(p)
  }
} else {
  cat("\n\n::: {.panel-tabset}\n\n")
  batch_list <- unique(amplicons_negative$Batch)
  for (batch in batch_list) {
    cat("## ", batch, "\n\n")
    batch_data <- amplicons_negative %>% filter(Batch == batch)
    total_negatives <- n_distinct(batch_data$negative)
    batch_data_nonzero <- batch_data %>% filter(!is.na(OutputPostprocessing) & OutputPostprocessing > 0)
    plotted_negatives <- n_distinct(batch_data_nonzero$negative)
    zero_read_negatives <- total_negatives - plotted_negatives

    if (nrow(batch_data_nonzero) == 0) {
      cat("Skipping batch:\n", batch, "No reads present in any negative controls in this batch.\n\n")
      next
    }

    cat(sprintf("Plotting reads for %d/%d negative controls (%d had no reads)\n\n", plotted_negatives, total_negatives, zero_read_negatives))
    n_controls_in_batch <- plotted_negatives
    batch_height <- min(max_height, max(min_height, ceiling(n_controls_in_batch / ncol_layout) * base_height_per_row))
    knitr::opts_chunk$set(fig.width = base_width, fig.height = batch_height)
    p <- plot_negative_control_histogram(batch_data_nonzero, reaction_colours, batch)
    if (!is.null(p)) print(p)
    cat("\n\n")
  }
  cat(":::\n\n")
}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

```{r negative-control-table}
cat("Summary of negative control targets with reads above threshold (", negative_control_read_threshold, ") \n")
neg_control_information <- amplicons_negative %>%
  filter(OutputPostprocessing > negative_control_read_threshold) %>%
  select("sample_name", "target_name", "OutputPostprocessing")

if (nrow(neg_control_information) == 0) {
  cat("No negative controls had targets with reads over threshold.\n")
}
head(neg_control_information)
```

#### Aggregated Reads per Target

```{r negative-control-aggregated}
summary_data <- amplicons_negative %>%
  group_by(reaction, target_name) %>%
  summarise(sum_reads = sum(reads, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    Chromosome = sub("-.*", "", target_name),
    Start = as.numeric(sub(".*-(\\d+)-.*", "\\1", target_name))
  ) %>%
  arrange(Chromosome, Start) %>%
  mutate(target_name = factor(target_name, levels = unique(target_name))) %>%
  mutate(reaction = factor(reaction)) %>%
  rowid_to_column("RowID")

summary_data <- amplicons_negative %>%
  filter(reads > 0) %>%
  group_by(target_name) %>%
  summarise(samples_with_reads = paste(unique(sample_name), collapse = ", "), .groups = "drop") %>%
  right_join(summary_data, by = "target_name")

nth_label <- 20
summary_data <- summary_data %>%
  mutate(Locus_label = ifelse(RowID %% nth_label == 1, paste0(Chromosome, ":", Start), ""))

ggplot(summary_data, aes(
  x = target_name, y = sum_reads, fill = reaction,
  text = paste("Target: ", target_name, "<br>Total Reads: ", sum_reads, "<br>Controls with Reads: ", samples_with_reads)
)) +
  geom_bar(stat = "identity", position = "dodge") +
  xlab("Target (Chromosome:Start Coordinate)") +
  ylab("Sum of Reads Across Controls") +
  ggtitle("Total Reads per Target Across Negative Controls") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 8),
    axis.title = element_text(size = 12),
    panel.grid = element_blank(),
    panel.background = element_blank(),
    legend.position = "right"
  ) +
  scale_x_discrete(labels = summary_data$Locus_label) +
  scale_fill_manual(values = reaction_colours)
```

## Successful Amplification

### Parasitemia vs. Number of Successfully Amplified Targets

```{r parasitemia-plots}
#| results: asis
generate_parasitemia_by_amplification_success_plot <- function(summary_samples_for_batch, threshold, sample_colours, batch_name) {
  ggplot(data = summary_samples_for_batch) +
    geom_point(aes(x = Parasitemia + 0.9, y = prop_good_loci, color = SampleType), shape = 1, alpha = 0.8, stroke = 1) +
    scale_x_log10() +
    ylim(0, 1) +
    facet_grid(rows = vars(reaction), scale = "free_y") +
    ylab(paste0("Amplicons with >", threshold, " reads")) +
    xlab("Parasitemia (log10)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle(paste0(batch_name, " Amplicons with `good` read depth")) +
    scale_color_manual(values = sample_colours)
}

generate_parasitemia_by_amplification_success_plots <- function(summary_samples, threshold, sample_colours) {
  plots <- list()
  for (b in unique(summary_samples$Batch)) {
    summary_samples_batch <- summary_samples %>% filter(Batch == b)
    plots[[b]] <- generate_parasitemia_by_amplification_success_plot(summary_samples_batch, read_threshold, sample_colours, b)
  }
  plots
}

plots <- generate_parasitemia_by_amplification_success_plots(summary_samples, read_threshold, sample_colours)
cat(create_tabsetOfGgplotObjects(plots))
```

### Sample Reads vs. Number of Successfully Amplified Targets

```{r reads-plots}
#| results: asis
generate_reads_by_amplification_success_plot <- function(summary_samples_for_batch, amplification_threshold,
                                                           sample_colours, reprep_threshold, repool_threshold, batch_name) {
  ggplot(data = summary_samples_for_batch) +
    geom_point(aes(x = reads_per_sample + 0.9, y = prop_good_loci, color = SampleType), shape = 1, alpha = 0.8, stroke = 1) +
    scale_x_log10() +
    scale_y_continuous(limits = c(0, 1)) +
    facet_grid(rows = vars(reaction), scale = "free_y") +
    ylab(paste0("Amplicons with >", amplification_threshold, " reads")) +
    xlab("Total Reads for Sample") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle(paste0(batch_name, " Targets that Amplified Successfully")) +
    scale_color_manual(values = sample_colours) +
    geom_hline(yintercept = reprep_threshold, linetype = "dashed") +
    geom_hline(yintercept = repool_threshold, linetype = "dashed") +
    annotate("text", x = min(summary_samples_for_batch$reads_per_sample, na.rm = TRUE), y = reprep_threshold, label = "Reprep", vjust = 1, hjust = 0) +
    annotate("text", x = min(summary_samples_for_batch$reads_per_sample, na.rm = TRUE), y = repool_threshold, label = "Repool", vjust = 1, hjust = 0)
}

generate_reads_by_amplification_success_plots <- function(summary_samples, read_threshold, sample_colours, reprep_threshold, repool_threshold) {
  plots <- list()
  for (b in unique(summary_samples$Batch)) {
    summary_samples_batch <- summary_samples %>% filter(Batch == b)
    plots[[b]] <- generate_reads_by_amplification_success_plot(summary_samples_batch, read_threshold, sample_colours, reprep_threshold, repool_threshold, b)
  }
  plots
}

read_plots <- generate_reads_by_amplification_success_plots(summary_samples, read_threshold, sample_colours, reprep_threshold, repool_threshold)
cat(create_tabsetOfGgplotObjects(read_plots))
```

## Per Sample QC Outcome Summary

```{r reprep-repool-summary}
generate_reprep_repool_table <- function(summary_samples, reprep_threshold, repool_threshold) {
  summary_samples %>%
    filter(SampleType != "negative") %>%
    mutate(
      status = case_when(
        prop_good_loci < reprep_threshold ~ "reprep",
        prop_good_loci < repool_threshold ~ "repool",
        TRUE ~ "pass"
      ),
      reason = case_when(
        prop_good_loci < reprep_threshold ~ paste("< reprep threshold ", reprep_threshold),
        prop_good_loci < repool_threshold ~ paste("< repool threshold ", repool_threshold),
        TRUE ~ "NA"
      )
    ) %>%
    arrange(desc(status)) %>%
    select(sample_name, Batch, reaction, status, reason, reads_per_reaction, prop_good_loci)
}

fill_missing_data <- function(reprep_repool_summary, manifest) {
  unique_reactions <- unique(reprep_repool_summary$reaction)

  missing_samples <- manifest %>%
    filter(!(sample_name %in% reprep_repool_summary$sample_name)) %>%
    select(sample_name, Batch)

  missing_sample_rows <- missing_samples %>%
    crossing(reaction = unique_reactions) %>%
    mutate(status = "reprep", reason = "no data for sample", reads_per_reaction = 0, prop_good_loci = 0)

  missing_reaction_rows <- reprep_repool_summary %>%
    select(sample_name, Batch) %>%
    distinct() %>%
    crossing(reaction = unique_reactions) %>%
    anti_join(reprep_repool_summary, by = c("sample_name", "Batch", "reaction")) %>%
    mutate(status = "reprep", reason = "no data for reaction", reads_per_reaction = 0, prop_good_loci = 0)

  bind_rows(reprep_repool_summary, missing_sample_rows, missing_reaction_rows)
}

reprep_repool_summary <- generate_reprep_repool_table(summary_samples, reprep_threshold, repool_threshold)
reprep_repool_summary <- fill_missing_data(reprep_repool_summary, manifest)
reprep_repool_summary %>% filter(status != "pass")

repreps_count <- reprep_repool_summary %>% filter(status == "reprep", !grepl("Control", sample_name)) %>% distinct(sample_name) %>% nrow()
repools_count <- reprep_repool_summary %>% filter(status == "repool", !grepl("Control", sample_name)) %>% distinct(sample_name) %>% nrow()
passedQC_count <- reprep_repool_summary %>% filter(status == "pass", !grepl("Control", sample_name)) %>% distinct(sample_name) %>% nrow()

total_samples <- repreps_count + repools_count + passedQC_count
if (total_samples > 0) {
  percentage_pass <- (passedQC_count / total_samples) * 100
  cat("Overall QC Summary:\n")
  cat("  Reprep:", repreps_count, "\n")
  cat("  Repool:", repools_count, "\n")
  cat("  Pass:", passedQC_count, "\n")
  cat("  Pass percentage:", round(percentage_pass, 1), "%\n")
}
```

## Amplification Plate Maps

### Reads Heatmap

```{r plate-heatmap-fn}
plot_plate_with_feature <- function(quadrants_batch, sample_colours, batch_name, fill_param, scale_midpoint, scale_label) {
  ggplot(quadrants_batch) +
    geom_tile(aes(x = Column, y = y, fill = !!rlang::parse_expr(fill_param)), color = NA, width = 0.95, height = 0.95) +
    geom_rect(
      data = quadrants_batch,
      aes(xmin = Column - 0.45, xmax = Column + 0.45, ymin = y - 0.45, ymax = y + 0.45, color = SampleType),
      fill = NA, linewidth = 1.3
    ) +
    facet_grid(reaction ~ Batch) +
    scale_fill_gradient2(low = "black", mid = "darkorange4", high = "darkorange", midpoint = scale_midpoint, name = scale_label) +
    scale_color_manual(values = sample_colours, na.translate = TRUE) +
    scale_y_continuous(breaks = quadrants_batch$y, labels = quadrants_batch$Row) +
    scale_x_continuous(breaks = 1:12, labels = as.character(1:12)) +
    xlab("Column") +
    ylab("Row") +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 12),
      strip.text = element_text(size = 14, face = "bold"),
      panel.grid = element_blank(),
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      legend.key = element_rect(colour = "black")
    ) +
    coord_fixed(ratio = 0.75)
}

plot_plate_heat_maps <- function(sample_data, quadrants, sample_colours, fill_param, scale_midpoint, scale_label) {
  plate_maps <- list()
  quadrants_by_reaction <- crossing(summary_samples %>% select(reaction) %>% distinct(), quadrants)
  merged_data <- merge(summary_samples, quadrants_by_reaction, by = c("Batch", "Row", "Column", "SampleType", "reaction"), all.y = TRUE)
  merged_data$SampleType[is.na(merged_data$SampleType)] <- "empty"

  for (b in unique(merged_data$Batch)) {
    merged_data_batch <- merged_data %>% filter(Batch == b)
    plate_maps[[b]] <- plot_plate_with_feature(merged_data_batch, sample_colours, b, fill_param, scale_midpoint, scale_label)
  }
  plate_maps
}
```

```{r reads-heatmap}
#| results: asis
plate_maps <- plot_plate_heat_maps(summary_samples, quadrants, sample_colours, "log10(reads_per_reaction + 0.1)", 2, "log10(reads_per_reaction)")
cat(create_tabsetOfGgplotObjects(plate_maps))
```

### Successful Amplification Heatmap

```{r amplification-heatmap}
#| results: asis
plate_maps <- plot_plate_heat_maps(summary_samples, quadrants, sample_colours, "prop_good_loci", 0.5, "prop_good_loci")
cat(create_tabsetOfGgplotObjects(plate_maps))
```

## QC Summary by Batch

Overall QC pass rate by batch, with stacked bars representing the percentage of samples passing,
requiring repool, or requiring reprep.

```{r qc-by-batch}
qc_summary_by_batch <- reprep_repool_summary %>%
  filter(!grepl("Control", sample_name)) %>%
  distinct(sample_name, Batch, status) %>%
  group_by(Batch, status) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(Batch) %>%
  mutate(total = sum(count), percentage = (count / total) * 100) %>%
  ungroup()

qc_colors <- c("pass" = "Gray", "repool" = "skyblue", "reprep" = "orangered2")

ggplot(qc_summary_by_batch, aes(x = Batch, y = percentage, fill = status)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3) +
  geom_text(
    aes(label = ifelse(percentage > 5, paste0(round(percentage, 1), "%"), "")),
    position = position_stack(vjust = 0.5), size = 5, fontface = "bold", color = "black"
  ) +
  scale_fill_manual(values = qc_colors, name = "QC Status", breaks = c("pass", "repool", "reprep"), labels = c("Pass", "Repool", "Reprep")) +
  labs(title = "QC Status by Batch", x = "Batch", y = "Percentage of Samples (%)") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "top",
    panel.grid.major.x = element_blank()
  ) +
  ylim(0, 100)

qc_summary_table <- reprep_repool_summary %>%
  filter(!grepl("Control", sample_name)) %>%
  distinct(sample_name, Batch, status) %>%
  group_by(Batch, status) %>%
  summarize(count = n(), .groups = "drop") %>%
  complete(Batch, status = c("pass", "repool", "reprep"), fill = list(count = 0)) %>%
  pivot_wider(names_from = status, values_from = count, values_fill = 0) %>%
  mutate(total = pass + repool + reprep, pass_rate = round(100 * pass / total, 1)) %>%
  select(Batch, pass, repool, reprep, total, pass_rate)

qc_summary_table %>%
  rename(Pass = pass, Repool = repool, Reprep = reprep, Total = total, `Pass Rate (%)` = pass_rate)
```

## Save Output Files

```{r save-outputs}
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
  cat("Created output directory:", output_dir, "\n")
}
reprep_repool_summary_path <- file.path(output_dir, reprep_repool_summary_path)
polyclonal_information_path <- file.path(output_dir, polyclonal_information_path)
neg_control_information_path <- file.path(output_dir, neg_control_information_path)
samples_to_repool_path <- file.path(output_dir, samples_to_repool_path)
samples_to_reprep_path <- file.path(output_dir, samples_to_reprep_path)
missing_samples_report_path <- file.path(output_dir, missing_samples_report_path)
filtered_allele_data_path <- file.path(output_dir, filtered_allele_data_path)
filtered_collapsed_allele_data_path <- file.path(output_dir, filtered_collapsed_allele_data_path)
missing_targets_report_path <- file.path(output_dir, missing_targets_report_path)
filtered_resmarker_table_path <- file.path(output_dir, filtered_resmarker_table_path)
filtered_resmarker_microhaplotype_table_path <- file.path(output_dir, filtered_resmarker_microhaplotype_table_path)

write.csv(reprep_repool_summary, reprep_repool_summary_path, row.names = FALSE)
cat("Saved:", reprep_repool_summary_path, "\n")

write.csv(polyclonal_information, polyclonal_information_path, row.names = FALSE)
cat("Saved:", polyclonal_information_path, "\n")

write.csv(neg_control_information, neg_control_information_path, row.names = FALSE)
cat("Saved:", neg_control_information_path, "\n")

samples_to_repool <- reprep_repool_summary %>%
  filter(status == "repool") %>%
  select(sample_name, Batch, reaction, reads_per_reaction, prop_good_loci, reason) %>%
  arrange(Batch, sample_name, reaction)

samples_to_reprep <- reprep_repool_summary %>%
  filter(status == "reprep") %>%
  select(sample_name, Batch, reaction, reads_per_reaction, prop_good_loci, reason) %>%
  arrange(Batch, sample_name, reaction)

write.csv(samples_to_repool, samples_to_repool_path, row.names = FALSE)
cat("Saved:", samples_to_repool_path, "(", nrow(samples_to_repool), "samples needing repool)\n")

write.csv(samples_to_reprep, samples_to_reprep_path, row.names = FALSE)
cat("Saved:", samples_to_reprep_path, "(", nrow(samples_to_reprep), "samples needing reprep)\n")

# Read in raw data with no other filtering applied
allele_data_for_output <- read.csv(allele_data_path, sep = "\t")
collapsed_allele_data_for_output <- read.csv(collapsed_allele_data_path, sep = "\t")
resmarker_table <- read.csv(resmarker_table_path, sep = "\t")
resmarker_microhaplotype_table <- read.csv(resmarker_microhaplotype_table_path, sep = "\t")

allele_data_for_output <- allele_data_for_output %>% mutate(sample_name = standardise_name(sample_name))
collapsed_allele_data_for_output <- collapsed_allele_data_for_output %>% mutate(sample_name = standardise_name(sample_name))

samples_in_allele_data <- allele_data_for_output %>% distinct(sample_name) %>% pull(sample_name)

missing_samples <- manifest %>%
  filter(!(sample_name %in% samples_in_allele_data)) %>%
  filter(SampleType != "negative") %>%
  select(sample_name, Batch, SampleType, Row, Column, Parasitemia) %>%
  arrange(Batch, sample_name)

write.csv(missing_samples, missing_samples_report_path, row.names = FALSE)
cat("Saved:", missing_samples_report_path, "(", nrow(missing_samples), "samples in manifest not found in allele data)\n")

allele_data_for_output <- allele_data_for_output %>%
  group_by(sample_name, target_name) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()

collapsed_allele_data_for_output <- collapsed_allele_data_for_output %>%
  group_by(sample_name, target_name) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()

resmarker_table <- resmarker_table %>%
  group_by(sample_name, gene_id, gene, aa_position) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()

group_cols <- c("sample_name", "gene_id", "gene", "target_name", "mhap_aa_positions", "ref_mhap")
resmarker_microhaplotype_table <- resmarker_microhaplotype_table %>%
  group_by(across(all_of(group_cols))) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()

if (filter_wsaf_final_allele_table) {
  allele_data_filtered <- allele_data_for_output %>% filter(reads > read_filter & AlleleFreq > af_filter)
  collapsed_allele_data_filtered <- collapsed_allele_data_for_output %>% filter(reads > read_filter & AlleleFreq > af_filter)
  resmarker_table_filtered <- resmarker_table %>% filter(reads > read_filter & AlleleFreq > af_filter)
  resmarker_microhaplotype_table_filtered <- resmarker_microhaplotype_table %>% filter(reads > read_filter & AlleleFreq > af_filter)
  filter_msg <- paste0("Reads > ", read_filter, " AND WSAF > ", af_filter)
} else {
  allele_data_filtered <- allele_data_for_output %>% filter(reads > read_filter)
  collapsed_allele_data_filtered <- collapsed_allele_data_for_output %>% filter(reads > read_filter)
  resmarker_table_filtered <- resmarker_table %>% filter(reads > read_filter)
  resmarker_microhaplotype_table_filtered <- resmarker_microhaplotype_table %>% filter(reads > read_filter)
  filter_msg <- paste0("Reads > ", read_filter, " (WSAF filter disabled)")
}

write.table(allele_data_filtered[, !(names(allele_data_filtered) %in% c("AlleleFreq"))], filtered_allele_data_path, sep = "\t", row.names = FALSE, quote = FALSE)
cat("Saved:", filtered_allele_data_path, "\n")
write.table(collapsed_allele_data_filtered[, !(names(collapsed_allele_data_filtered) %in% c("AlleleFreq"))], filtered_collapsed_allele_data_path, sep = "\t", row.names = FALSE, quote = FALSE)
cat("Saved:", filtered_collapsed_allele_data_path, "\n")
write.table(resmarker_table_filtered[, !(names(resmarker_table_filtered) %in% c("AlleleFreq"))], filtered_resmarker_table_path, sep = "\t", row.names = FALSE, quote = FALSE)
cat("Saved:", filtered_resmarker_table_path, "\n")
write.table(resmarker_microhaplotype_table_filtered[, !(names(resmarker_microhaplotype_table_filtered) %in% c("AlleleFreq"))], filtered_resmarker_microhaplotype_table_path, sep = "\t", row.names = FALSE, quote = FALSE)
cat("Saved:", filtered_resmarker_microhaplotype_table_path, "\n")
cat("  (Includes ALL targets, including those excluded from QC calculations)\n")
cat("  Filtered by:", filter_msg, "\n")
```

### Missing Loci Report

```{r missing-loci}
loci_not_in_allele_data <- panel_information %>%
  anti_join(allele_data, by = "target_name") %>%
  select(target_name, pool, chrom)

write.csv(loci_not_in_allele_data, missing_targets_report_path, row.names = FALSE)
cat("Saved:", missing_targets_report_path, "\n")

if (nrow(loci_not_in_allele_data) > 0) {
  cat("\nMissing Targets Report\n\n")
  cat("The following targets are present in the pool file but missing from the allele data:\n\n")
  loci_not_in_allele_data %>%
    arrange(target_name) %>%
    mutate(No = row_number()) %>%
    select(No, everything())
} else {
  cat("\nAll targets in pool file were found in the data.\n")
}
```

## Report Summary

```{r report-summary}
cat("This unified QC report has processed data and generated the following standardized outputs:\n\n")
cat("1.", reprep_repool_summary_path, "- Combined reprep/repool classification for all samples\n")
cat("2.", samples_to_repool_path, "- Samples requiring repool\n")
cat("3.", samples_to_reprep_path, "- Samples requiring reprep\n")
cat("4.", polyclonal_information_path, "- Polyclonality information for positive controls\n")
cat("5.", neg_control_information_path, "- Contamination detected in negative controls\n")
cat("6.", filtered_allele_data_path, "- Filtered allele data meeting read/AF thresholds (includes all targets)\n")
cat("7.", missing_targets_report_path, "- Targets in pool file but absent from data (if applicable)\n")
cat("8.", missing_samples_report_path, "- Samples in manifest but not found in allele data\n\n")

cat("- Targets filtered for QC:", length(excluded_loci), "targets were excluded from QC calculations\n")
cat("- QC metrics: Reprep/repool classifications based on", dplyr::n_distinct(amplicon_coverage_qc$target_name), "QC-relevant targets\n")
cat("- Final output: allele_data_filtered.txt includes ALL targets (both QC and excluded) meeting thresholds\n")
if (exists("percentage_pass")) {
  cat("- Pass rate:", paste0(round(percentage_pass, 1), "%"), "\n")
} else {
  cat("- Pass rate: N/A\n")
}
```
